Stack Safety as a Security Property

### 2020-07-10 main.tex
read through 3

We assume we are given a machine language program, together with some auxiliary information
about its structure: the code footprint of each function and the locations of instructions
representing calls, returns, and function entry points.
Although this information would typically come from a compiler together with
the machine code, we do not
assume anything about their provenance.  We assume that the machine has a
program counter $\PCname$
and stack pointer $\rsp$\bcp{but not a frame pointer?!}; other details of the machine's ISA are unimportant.

To make our stack safety definitions precise, we make some straightforward
assumptions about
memory layout; our definitions could easily be changed to be parametric over
these assumptions, at the cost of some complexity in their statements.  Memory
consists of an instruction region and a stack that grows upwards. (For simplicity,
we do not consider the heap. Integrating a memory safety property for the
heap in the style of
\citet{DBLP:conf/post/AmorimHP18} is a natural extension that we discuss
in~\Cref{sec:relwork}.)  The stack has a conventional structure, with
contiguously allocated per-function frames.
The boundary between the frames of a caller and callee is indicated by the $\rsp$ register
at the time of the call.  The stack can be used to
pass arguments or return results in the usual way, by placing them
at the top of the caller's frame. Callers and callees can also communicate
via registers if they wish; indeed, our
definitions say nothing at all about registers (except $\PCname$ and
$\rsp$). \bcp{This implicitly amounts to choosing a caller-saves register
  convention, though, doesn't it?}

Our stack safety concepts depend critically on the notions of ``call'' and ``return,''
which are not fully explicit in machine code (the same instruction may be
used in some places as part of a call or return sequence and in other places
for other purposes).  We define a call to be
any instruction declared as such in the compiler-provided call structure.
If a call occurs at $\PCname = a$ and $\rsp = s$, then the \emph{corresponding
  return} has occurred when $\PCname = \nextPC(a)$ and $\rsp = s$. (Our definition
of well-bracketed control flow relates this definition of return to the program's
declared set of return points.)
Our property definitions make no other detailed assumptions about call or return sequences, although
some of the enforcement mechanisms we consider do.

@@

We model ordinary program state and policy-enforcement state separately to make it
easier to use well-established concepts from language-based security.
In particular, confidentiality has an elegant extensional characterization in terms of
\emph{noninterference}: a program preserves the confidentiality of certain data if
\emph{varying} that data would not change the program's observable
behavior~\citep{6234468}
In applying this idea to a system that incorporates policy enforcement, only
ordinary program data should be varied, while any associated policy
state should be left alone; our model makes this distinction obvious.
%That is, we compare the actual execution of the program with a \emph{hypothetical}
%execution where data has been varied.
We will also introduce a notion of {\em observational integrity} that relies on the idea of
\emph{rolling back} program state; again, our model makes it easy to describe
rollback of ordinary program data while leaving policy state alone.

@@

\paragraph*{Micropolicies}
%
Micropolicies are a flexible tag-based, hardware-accelerated
reference monitoring mechanism; they been applied to stack safety
enforcement by \citet{DBLP:conf/sp/RoesslerD18}.
Here the policy state consists of metadata tags (e.g., identifying stack frames)
attached to each value in memory or registers, and the policy step
function\bcp{undefined!}\leo{``partial policy step'' is explained informally above}
checks that each machine operation obeys a set of rules on tags
(e.g., that the current $\PCname$ tag matches the tag on the stack
location being accessed), and halts the machine if not.  Distinguishing
ordinary machine state and policy state is very natural in this setting.
%% Here the machine state is extended with tags paired with all components of the
%% system, and a monitor checks for each operation that the tags on the operands
%% obey a set of rules. One could treat tags directly as part of the machine
%% state, and build the rules into the step function. But local state
%% encapsulation properties as we frame them are not intuitive in their
%% interaction with tags or other additional enforcement state, so it is helpful
%% to separate tags for clarity. This also distinguishes a hardware fail-stop from
%% a software fail state.
%% \rb{TODO Similarly to above, should we consider tags as part of the machine or part of the
%%   policy? }

\paragraph*{Capability machines}
%
%Other policies are implemented in hardware or built on hardware mechanisms.
Capability machines such as CHERI \citep{Woodruff+14} extend conventional
architectures to support efficient and fine-grained control over memory
accesses, which can be used to implement security policies, including
the stack protection policies of
\citet{Skorstengaard+19,Skorstengaard+19b}\bcp{I thought these policies used
  linear capabilities?}\leo{Only the second one AFAIK. What would you
  change to characterize it more accurately?}\rb{StkPointers uses linear
  capabilities, its predecessor users local capabilities (and unlike
  StkPointers could be implemented in CHERI)}.
Capabilities package up pointers
(ordinary data) with base and bounds information (policy data),
and they require rewriting source code (e.g., to use capabilities in place
of ordinary pointers), so separating ordinary machine state from policy
state is delicate; we conjecture it can be done with some effort.

@@

\paragraph*{Software-only policies}
%
Some policies are implemented in software, by modifying code in the
compiler, for example to perform bounds checking~\citep{NagarakatteZMZ09} or
to insert stack canaries~\citep{Cowan+98}. Such approaches should
still satisfy our formal notion of stack safety with a trivial null
policy, but only for programs produced by the compiler.
%These approaches are represented as policies that
%accept\bcp{what does it mean for a policy to accept a state??} only initial machine states that are well-formed for the policy (due to
%being produced by the compiler.) The set of policy states is again the unit type
%and the step function constant. Property enforcement comes from the fact that
%only ill-formed states can induce a trace that violates the property.

@@

\Cref{fig:running-program} shows a high-level view of our running example,
written in C-like pseudo-code.  It consists of a {\tt main} function that
performs a call to some function {\tt f}, which in turn makes a nested call
to some function {\tt g} before returning, while performing some simple
arithmetic operations in the process. \Cref{fig:running-trace-b} shows an
execution trace of the corresponding machine instructions. These might have
been generated by compiling the high-level code, but it is only the behavior
of the machine instructions that we care about
here---\Cref{fig:running-trace-b} is our actual example.

@@

A sketch of the initial memory layout appears at the top of
\Cref{fig:running-trace-b}: the instructions for {\tt main} begin at 0,
and the (uninitialized)
stack begins at 100. Directly below the memory sketch, we summarize the parts
of the register file that are relevant at this point in the example.
%
The rest of \Cref{fig:running-trace-b} traces the execution of the
program, showing the machine state at call and return points.
Instructions are intended to
be similar to RISC-V, but are described in a (hopefully more readable) an
ad-hoc notation
rather than using assembler syntax. The $\jal$ (jump-and-link)
instruction used for calls atomically simultaneously the return address in a
register
and jumps to a specified new $\PCname$; the $\jalr$ (jump-and-link-via-register)
instruction used for returns jumps to the address stored in a register.
Register $r_0$ is hardwired to zero, so that, e.g., instruction 3 stores $0$
into address 101.

@@

>>>>

We assume we are given a machine language program, together with some auxiliary information
about its structure: the code footprint of each function and the locations of instructions
representing calls, returns, and function entry points.
Although this information would typically come from a compiler together with
the machine code, we do not
assume anything about their provenance.  We assume that the machine has a
program counter $\PCname$
and stack pointer $\rsp$\bcp{but not a frame pointer?!\apt{right; we're agnostic about that}}; other details of the machine's ISA are unimportant.

To make our stack safety definitions precise, we make some straightforward
assumptions about
memory layout; our definitions could easily be changed to be parametric over
these assumptions, at the cost of some complexity in their statements.  Memory
consists of an instruction region and a stack that grows upwards. (For simplicity,
we do not consider the heap. Integrating a memory safety property for the
heap in the style of
\citet{DBLP:conf/post/AmorimHP18} is a natural extension that we discuss
in~\Cref{sec:relwork}.)  The stack has a conventional structure, with
contiguously allocated per-function frames.
The boundary between the frames of a caller and callee is indicated by the $\rsp$ register
at the time of the call.  The stack can be used to
pass arguments or return results in the usual way, by placing them
at the top of the caller's frame. Callers and callees can also communicate
via registers if they wish; indeed, our 
definitions say nothing at all about registers (except $\PCname$ and
$\rsp$). \bcp{This implicitly amounts to choosing a caller-saves register
  convention, though, doesn't it?\apt{I think it is neutral; we could add separate caller register protection if we wanted.}}

Our stack safety concepts depend critically on the notions of ``call'' and ``return,''
which are not fully explicit in machine code (the same instruction opcode may be
used in some places as part of a call or return sequence and in other places
for other purposes).  We define a call to be
any instruction declared as such in the compiler-provided call structure.
If a call occurs at $\PCname = a$ and $\rsp = s$, then the \emph{corresponding
  return} has occurred when the $\PCname$ points to the next instruction after $a$ and $\rsp = s$. (Our definition
of well-bracketed control flow relates this definition of return to the program's
declared set of return points.)
Our property definitions make no other detailed assumptions about call or return sequences, although
some of the enforcement mechanisms we consider do.

@@

We model ordinary program state and policy-enforcement state separately to make it
easier to use well-established concepts from language-based security.
In particular, confidentiality has an elegant extensional characterization in terms of
\emph{noninterference}: a program preserves the confidentiality of certain data if
\emph{varying} that data would not change the program's observable
behavior~\citep{6234468}.
In applying this idea to a system that incorporates policy enforcement, only
ordinary program data should be varied, while any associated policy
state should be left alone; our model makes this distinction obvious.
%That is, we compare the actual execution of the program with a \emph{hypothetical}
%execution where data has been varied.
We will also introduce a notion of {\em observational integrity} that relies on the idea of
\emph{rolling back} program state; again, our model makes it easy to describe
rollback of ordinary program data while leaving policy state alone.

@@

\paragraph*{Micropolicies}
%
Micropolicies are a flexible tag-based, hardware-accelerated
reference monitoring mechanism; they been applied to stack safety
enforcement by \citet{DBLP:conf/sp/RoesslerD18}.
Here the policy state consists of metadata tags (e.g., identifying stack frames)
attached to each value in memory or registers, and the policy step
function checks that each machine operation obeys a set of rules on tags
(e.g., that the current $\PCname$ tag matches the tag on the stack
location being accessed), and halts the machine if not.  Distinguishing
ordinary machine state and policy state is very natural in this setting.
%% Here the machine state is extended with tags paired with all components of the
%% system, and a monitor checks for each operation that the tags on the operands
%% obey a set of rules. One could treat tags directly as part of the machine
%% state, and build the rules into the step function. But local state
%% encapsulation properties as we frame them are not intuitive in their
%% interaction with tags or other additional enforcement state, so it is helpful
%% to separate tags for clarity. This also distinguishes a hardware fail-stop from
%% a software fail state.
%% \rb{TODO Similarly to above, should we consider tags as part of the machine or part of the
%%   policy? }

\paragraph*{Capability machines}
%
%Other policies are implemented in hardware or built on hardware mechanisms.
Capability machines such as CHERI \citep{Woodruff+14} extend conventional
architectures to support efficient and fine-grained control over memory
accesses, which can be used to implement security policies, including
the stack protection policies of
\citet{Skorstengaard+19b} and (with an extension to support linear capabilities) \citet{Skorstengaard+19}.
%\bcp{I thought these policies used
%  linear capabilities?}\leo{Only the second one AFAIK. What would you
%  change to characterize it more accurately?}\rb{StkPointers uses linear
%  capabilities, its predecessor users local capabilities (and unlike
%  StkPointers could be implemented in CHERI)}.
Capabilities package up pointers
(ordinary data) with base and bounds information (policy data),
and they require rewriting source code (e.g., to use capabilities in place
of ordinary pointers), so separating ordinary machine state from policy
state is delicate; we conjecture it can be done with some effort.

@@

\paragraph*{Software-only policies}
%
Some policies are implemented in software, by modifying code in the
compiler, for example to perform bounds checking~\citep{NagarakatteZMZ09} or
to insert stack canaries~\citep{Cowan+98}. Such approaches should
still satisfy our formal notion of stack safety \apt{Stack canaries certainly would not!} with a trivial null
policy, but only for programs produced by the compiler.\apt{This still seems wrong to me. The policy part needs somehow to describe everything that isn't varied or rolled back.}
%These approaches are represented as policies that
%accept\bcp{what does it mean for a policy to accept a state??} only initial machine states that are well-formed for the policy (due to
%being produced by the compiler.) The set of policy states is again the unit type
%and the step function constant. Property enforcement comes from the fact that
%only ill-formed states can induce a trace that violates the property.

@@

\Cref{fig:running-program} shows a high-level view of a running example,
written in C-like pseudo-code.  It consists of a {\tt main} function that
performs a call to some function {\tt f}, which in turn makes a nested call
to some function {\tt g} before returning, while performing some simple
arithmetic operations in the process. \Cref{fig:running-trace-b} shows an
execution trace of the corresponding machine instructions. These might have
been generated from the high-level code by a (rather simple-minded) compiler,
but it is only the behavior 
of the machine instructions that we care about
here---\Cref{fig:running-trace-b} is our actual example.

@@

A sketch of the initial memory layout appears at the top of
\Cref{fig:running-trace-b}: the instructions for {\tt main} begin at 0,
and the (uninitialized)
stack begins at 100. Directly below the memory sketch, we summarize the parts
of the register file that are relevant at this point in the example.
%
The rest of \Cref{fig:running-trace-b} traces the execution of the
program, showing the machine state at call and return points.
Instructions are intended to
be similar to RISC-V, but are described in a (hopefully more readable)
ad-hoc notation
rather than using assembler syntax. The $\jal$ (jump-and-link)
instruction used for calls atomically simultaneously the return address in a
register
and jumps to a specified new $\PCname$; the $\jalr$ (jump-and-link-via-register)
instruction used for returns jumps to the address stored in a register.
Register $r_0$ is hardwired to zero, so that, e.g., instruction 3 stores $0$
into address 101.

The code itself is straightforward, and should be largely self-explanatory, but
a few points deserve clarification. Functions return their result value on the
stack rather than in a register.  Stack frames are built in two parts:
space for local variables and the return address is allocated and initialized at function entry, and deallocated at function exit; space for arguments and return values is allocated and initialized immediately prior to each function call and deallocated on return. The return value slot is always initialized
because some of our enforcement mechanisms will rely on this.
\apt{this is bogus, but oh well}
The return address is spilled to the stack even by leaf functions such as
{\tt h} that could just leave it in $\rra$.

@@

\apt{It would be good to relate this back to the running example. This is
  complicated because the example passes arguments/returns, and it builds
  and initializes this part of the frame separately.}
